My wonderful paper

### 2024-12-10 _manuscript/_tex/index.tex
render pdf

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Roger
D.
Peng. Email: \href{mailto:roger.peng@austin.utexas.edu}{\nolinkurl{roger.peng@austin.utexas.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Texas, United States
\begin{abstract}
In data analysis, unexpected results often prompt researchers to revisit
their procedures to identify potential issues. While experienced
researchers can typically diagnose problems by checking a few key
assumptions, others may struggle to identify the root causes. These
checked assumptions, or expectations, are typically informal, difficult
to trace, and rarely discussed in publications. In this paper, we
formalize these informal assumptions by framing them as binary
\emph{analysis validation check}. We then introduce a procedure to
quantify how violations of these checks may lead to unexpected results
in the analysis. The procedure relies on simulations of the original
data and evaluates both accuracy and redundancy. Accuracy is calculated
through binary classification metric, while redundancy is measured using
mutual information. We demonstrate this approach with a toy example on
the fitness step count data and a generalized linear model exmaple
examining the effect of PM10 on mortality.
\end{abstract}

@@

\begin{itemize}
\tightlist
\item
  TODO: emphasize trustworthy data science since it is the theme of the
  special issue
\end{itemize}

In data analysis, experienced researchers often rely on their prior
knowledge or domain expertise to quickly assess whether results align
with their expectations. When a result falls outside of this interval,
it prompts the researchers to investigate backwards on the data quality,
the analysis steps, or the assumptions made during the analysis process.
This mental process of where to diagnose unexpected outcomes is often
difficult to trace and discuss in publications. As a result, readers are
typically presented with the final outcomes of the analysis cycle where
the results and expectations are aligned -- achieved either by refining
the analysis or updating the expectations based on statistical evidence
\citep{grolemund_cognitive_2014}. These missing pieces of information
provides little guidance for diagnosing issues in the analysis when the
same methodology is applied to a new dataset that produces different
outcomes. Similarly, when researchers with different background
knowledge it becomes unclear whether discrepancies arise from differing
expectations or from the use of statistical techniques.

One might gain insight into analysts' thought process by speaking with
them directly or watching screencast videos they produce, such as,
TidyTuesday screencast videos. However, direct conversation are not
scalable and may not be feasible while creating educational videos
requires significant effort from the researchers. Ideally, there could
be a way to make these expectations explicit and accessible to others.
Even better, if the encoding is machine-readable, we could analyze these
expectations and learn from the analysis -- whether the checks also
apply to other researchers analyzing new data in the same context,
whether they reflect common practices in the field, or whether they are
specific to the data or analysis at hand.

In this paper, we conceptualize these internal expectations and
assumptions as \emph{analysis validation checks}, which allows us to
examine these assumptions made during analysis to diagnose unexpected
outcomes. We then introduce a procedure that provides a quantitative
measure on how violations in a tree of checks, derived from individual
checks, will lead to an unexpected result. The procedure, based on
simulations of the original data, calculates accuracy and redundancy.
Accuracy is determined using binary classification metrics, precision
and recall, from a logic regression fit while redundancy is measured
using mutual information. The proposed workflow offers a numerical
guarantee that the analysis will produce the expected results, assuming
the data generating mechanism holds. {[}something about trustworthy data
science{]}

The rest of the paper is organized as follows:
Section~\ref{sec-lit-review} reviews the concept of diagnosing
unexpected outcomes and data quality checks. Section~\ref{sec-plan}
introduces the concept of analysis validation check, illustrated with a
toy example on fitness step count. Section~\ref{sec-method} describes
the procedure that quantify how checks combined in logical operators can
predict unexpected outcomes. Section~\ref{sec-pm10-mortality} applies
this procedure to examine the effect of PM10 on mortality in public
health in a generalized linear model. Section~\ref{sec-discussion}
discusses a few key considerations and Section~\ref{sec-conclusion}
concludes the paper.

\section{Literature review}\label{sec-lit-review}

@@

\begin{itemize}
\item
  They shape how we interpret the results and assess whether they are
  consistent with existing knowledge or indicate the need for updates.
  \citep{grolemund_cognitive_2014}
\item
  \citep{peng_diagnosing_2021} describes three pedagogical exercises of
  introducing diagnosing unexpected outcome into a classroom setting.
\item
  what if the expectation is ``wrong''
\end{itemize}

@@

A substantial body of literature has addressed the definition of data
quality \citep[more]{8642813} and developed frameworks, that includes
dimensions, attributes, and measures to evaluate and improve data
quality
\citep{cai2015challenges, wang1996beyond, 6204995, woodall2014classification}.
These frameworks are often used information system and database
management and support business decision-making in the industry. For
research purposes, high-quality data ensures the credibility of
scientific findings and supports reproducibility and reusability in
future studies {[}ref{]}. With the growing prevalence of open data in
scientific research, the consumers or users of the data typically are no
longer the data producers or collectors who have the full knowledge of
data in hand, prompting more interest towards the data quality checks in
the data analysis process. In R, there are some packages, like
\texttt{skimr} \citep{skimr} and \texttt{dataMaid} \citep{dataMaid},
provides basic data screening and reporting, while another class of
packages, e.g.~\texttt{assertr} \citep{assertr}, \texttt{validate}
\citep{validate}, and \texttt{pointblank} \citep{pointblank} focuses on
providing data validation tools, allowing users to define customized
data quality checks based on the applications.

@@

\subsection{Framing expectations as
checks}\label{framing-expectations-as-checks}

Expectations represent our beliefs about certain aspects of the
analysis, independent of the analysis itself. When outcomes deviate from
these expectations, analysts often revisit the analysis cycle to
identify potential issues, refine methods, or revise assumptions.
Experienced analysts can typically identify issues quickly and correct
them on the spot, but they often do so without discussing the underlying
reasoning, making it harder for less experienced researchers to learn
and master these skills. Here, we introduce the concept of
\emph{analysis validation checks}, which frames these expectations or
assumptions as explicit checks that returns a TRUE or FALSE output given
the data analyzed at hand. Inspired by data validation checks
\citep{validate}, where ensure datasets meet expected format and
quality, analysis validation checks reverse the approach: they validate
the assumptions about the data necessary for the analysis to produce the
reported results. This allows the concept to encompass a broad range of
checks, such as data quality (i.e.~missing data, how the data is
structured), data distribution and outliers, bivariate and multivariate
relationships between variables, and contextual information. These
externalized checks provide insights into analysts' thoughts process
process and offer several benefits:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  help students and junior researchers develop skills to diagnose
  unexpected analysis results,
\item
  serve as clear checkpoints to support replication or application of
  methods to (new) data,
\item
  align assumptions among researchers from different domain backgrounds,
  and
\item
  improve analysis transparency, reproducibility, and trustworthiness.
\end{enumerate}

\subsection{A toy example}\label{a-toy-example}

Consider a 30-day step count experiment in public health. Testers are
instructed to walk at least 8,000 steps each day, with an expected
average of 9,000 steps, tracked by a step counter app. While there may
be occasional ``low'' days due to factors such as forgetting to wear
their watch or unfavorable weather condition limiting outdoor
activities, there are also ``high'' days after an outdoor hike or
intense workout. After 30 days, you review the data and expect no more
than five days with step counts falling below the 8,000-step threshold
{[}I think we need a bit more scenario to justify we expect no more than
5 days, say it needs to have at least 25 days for it to be valid for the
next stage research??{]}.

To simulation this data, three normal distributions are used for the
daily step counts: \(\mathcal{N}(4000, 200)\) for low days,
\(\mathcal{N}(12000, 200)\) for high days, and
\(\mathcal{N}(9000, 300)\) for typical days. The number of low and high
days can be simulated from a Poisson distribution with \(\lambda = 4\).
Figure~\ref{fig-step-count} displays the number of days with fewer than
8,000 steps across 300 simulated 30-day periods.

In this scenario, the outcome is that the number of days with a step
count below 8,000 being no more than five. To diagnose potential reasons
why this outcome expectation might fail, we can establish a few analysis
validation checks. For example, if the average step count is too low,
this may suggest there are too many low days, potentially lead to the
unexpected outcome. Similarly, we can also check the quantile of the
step count, if more than a third of the days fall below 8,000, this
could indicate an excess of low-count days. Additionally, we may may
expect the standard deviation of the step count not to be overly large.
These considerations yield the following three unit tests as plan
expectations:

\begin{itemize}
\tightlist
\item
  test1: the test fails if the mean step count is below 8,200
\item
  test2: the test fails if the 30th percentile of the step counts is
  below 8,200
\item
  test3: the test fails if the standard deviation of the step counts
  exceeds 2,500.
\end{itemize}

@@

@@

\subsection{A workflow to assess the quality of unit
tests}\label{a-workflow-to-assess-the-quality-of-unit-tests}

While some checks are crucial and can lead to unexpected outcomes,
others may be tangential to the root cause. In this section, we propose
a procedure to measure the violation of checks, combined using logical
operators, contribute to an unexpected outcome. A small set of
independent checks is considered effective if it translates to
unexpected outcomes and Figure~\ref{fig-metric-calc} provides an
overview of the calculation.

@@

From the simulated data, the accuracy branch refers to a set of checks'
ability to accurately detect unexpected outcomes while minimizing false
positives and false negatives. While a false positive can raise caution
or skepticism on the data, both false positive and false negative
suggest the checks are over or lack sensitivity to unexpected outcomes.
To incorporate the effect of multiple checks on the outcome, a logic
regression model \citep{ruczinski_logic_2003} is fitted to the checks
and whether the outcome is unexpected. Originally developed for SNP
microarray data, logic regression constructs Boolean combinations of
binary variables, in a tree structure, to predict both binary and
numerical outcomes. Compared to other tree-based methods for
binary-binary prediction, the Boolean combinations from the logic
regression model produce a tree structure that can be directly
interpreted as the possible combination of checks leading to an
unexpected outcome, without the need to invert the tree as required in
classic tree-based methods. The model is then used to predict the
outcome expectation based on the checks, and the prediction is compared
to the actual outcome expectation to calculate the precision and recall
of the checks.

While tests may score high on accuracy, they may be less effective at
explaining the reasons behind unexpected results. This could happen if a
set of tests are all tangentially related to the cause of the unexpected
results, but none addressing the root cause. It may also occur if the
tests are correlated with one another, leading to redundancy.

@@

To combine precision, recall, and independence into a single metric,
mathematical means, such as arithmetic mean, harmonic mean, and
quadratic mean can be used. The differences among these means are
minimal when the three metrics are similar. However, as the differences
among the metrics increases, the harmonic mean tends to produce the
smallest overall score, as it penalizes low values, while the quadratic
mean tends to produce the largest score by rewarding higher values more.
For simple interpretation of the score, the arithmetic mean is
preferred, while in applications where the difference between precision,
recall, and independence need to be penalized or rewarded more, the
harmonic and quadratic mean should be considered.

\subsection{Toy example revisited}\label{toy-example-revisited}

@@

@@

In the step count example, the logic regression model is fitted to the
three unit tests to generate the prediction for calculating precision
and recall. Figure~\ref{fig-logic-reg} shows the chosen logic regression
model: a combination of test 1 and test 3 with an OR operator and
Table~\ref{tbl-logic-reg} presents the calculated precision, recall, and
independence for the three individual tests and the combined test rule
(test1 OR test3) from the logic regression. We also include the metric
calculated from fitting a regression tree model to the data to compare
the performance of the logic regression model. The harmonic and
arithmetic means are included to combine the three measures. The results
show that the two tests produced by the logic regression can accurately
predict 82.1\% of all \emph{predicted unexpected cases}, and 83.6\%
cases of all \emph{actual unexpected results} in the simulation data.
The regression tree produces a similar model of first split on test1 and
then split on test3, and results in the same accuracy and overall score
as the logic regression model.

@@

Consider a generalized linear model (GLM) to study the effect of PM10 on
mortality {[}TODO: provide more context of the mortality-PM10 study{]}.
Analysts may expect a PM10 coefficient between {[}0, 0.005{]} after
considering the temperature confounding {[}TODO: reference{]}. The
example demonstrates how the procedure presented in
Section~\ref{sec-method} can be used to select cutoff values in the
checks to diagnose an unexpected PM10 coefficient from the generalized
linear model.

This PM10 coefficient expectation can be framed into a check that fails,
labelled as 1, if the PM10 coefficient is outside the range {[}0,
0.005{]} and 0 if it is within. Multiple factors can affect the PM10
coefficient, such as the sample size, the strength of the correlation
between mortality and PM10, and the strength of the correlation between
mortality and temperature. Analysts may expect a reasonable sample size
to ensure the reliability of the coefficient estimate. Outliers in the
three variables can also leverage the coefficient. While these are
possible factors that could affect the analysis result, it is not clear
the cutoff values for these checks to determine a failure. Here we
create a list of checks in Table~\ref{tbl-checks} with varied cutoff
values for each:

@@

To generate replicate of the data, we first simulate the correlation
matrix of the three variables in a grid and then use a Gaussian copula
to generate a multivariate normal distribution based on the specified
correlation matrix and sample size. The multivariate normal distribution
is transformed using the normal CDF before the inverse CDF of the
assumed distributions of the three variables is applied. To determine
the appropriate distribution of each variable, various distributions are
fitted and compared. This includes poisson and negative binomial for
mortality; gamma, log-normal, exponential, weibull, and normal for pm10
and temperature; and beta for pm10 after rescaling the data to
\([0-1]\). AIC is used to determine the best distribution fit for each
variable with qq plot presented in Figure~\ref{fig-dist-fit} to evaluate
the fit. AIC suggests a negative binomial distribution for mortality, a
beta distribution for PM10 (multiple by 100 to obtain the original
scale), and a Weibull distribution for temperature. To include the
effect of outlier, we add a single outlier to the data for both variable
mortality and PM10.

@@

@@

A logic regression is fitted to predict whether the PM10 coefficient is
unexpected using the checks listed in Table~\ref{tbl-checks} and
Figure~\ref{fig-linear-reg-tree} shows the logic regression tree from
the fitted model. Precision, recall, and independence score, along with
their harmonic and arithmetic mean are calculated in
Table~\ref{tbl-linear-reg}. The logic regression model picks up the
following cutoff value for each type of check:

\begin{itemize}
\tightlist
\item
  sample size \emph{larger than} 200
\item
  mortality-temperature correlation greater than -0.35
\item
  mortality-PM10 correlation less than -0.04
\item
  mortality contain outliers that are detected by the univariate outlier
  detection
\end{itemize}

@@

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Roger
D.
Peng. Email: \href{mailto:roger.peng@austin.utexas.edu}{\nolinkurl{roger.peng@austin.utexas.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Texas, United States
\begin{abstract}
In data analysis, unexpected results often prompt researchers to revisit
their procedures to identify potential issues. While experienced
researchers can often quickly diagnose problems by checking a few key
assumptions, others may struggle to identify the root causes. These
checked assumptions, or expectations, are typically informal, difficult
to trace, and rarely discussed in publications. In this paper, we
formalize these informal assumptions by framing them as binary
\emph{analysis validation checks}. We then introduce a procedure to
quantify how violations of these checks may lead to unexpected results
in the analysis. The procedure relies on simulations of the original
data and evaluates both accuracy and redundancy. Accuracy is calculated
through a binary classification metric, while redundancy is measured
using mutual information. We demonstrate this approach with a toy
example based on fitness step count data and a generalized linear model
example examining the effect of particulate matter air pollution on
daily mortality.
\end{abstract}

@@

In data analysis, experienced researchers often rely on their prior
knowledge or domain expertise to quickly assess whether results align
with their expectations. When a result falls outside of this interval,
it prompts the researchers to investigate backwards on the data quality,
the analysis steps, or the assumptions made during the analysis process.
This mental process of where to diagnose unexpected outcomes is often
difficult to trace and discuss in publications. As a result, readers are
typically presented with the final outcomes of the analysis cycle where
the results and expectations are aligned, achieved either by refining
the analysis or updating the expectations based on statistical evidence
\citep{grolemund_cognitive_2014}. These missing pieces of information
provides little guidance for diagnosing issues in the analysis when the
same methodology is applied to a new dataset that produces different
outcomes. Similarly, when researchers with different background
knowledge view results that they find to be unexpected, it becomes
unclear whether discrepancies arise from differing expectations or from
the use of statistical techniques.

One might gain insight into analysts' thought processes by speaking with
them directly or watching them work via screencast videos they produce,
such as, TidyTuesday screencast videos or think-aloud type studies
\citep[e.g.][]{gu2024data}. However, direct observation of analysis is
not scalable and may not always be feasible; creating educational
screencast videos requires significant effort from the researchers.
Ideally, there could be a way to make expectations about data analysis
explicit and accessible to others. Even better, if the encoding were
machine-readable, we could analyze these expectations and learn from the
analysis itself. For example, we could answer questions about whether
the checks also apply to other researchers analyzing new data in the
same context, whether they reflect common practices in the field, or
whether they are specific to the data or analysis at hand.

The externalization of the data analysis process is a practice that has
potential to improve the trustworthiness of analyses in general and the
trustworthiness of subsequent products, such as machine learning models,
that may be built on such analyses. While publication of analysis code
and data is now a common a requirement for the sake of reproducibility
\citep{peng2011reproducible}, the publication code alone is often
insufficient for understanding the thought process behind an analysis.
Code corresponding to published analyses often reflect the final
decisions made about analysis and do not reveal the decisions or
assumptions made about the data processing. Thus, a reader looking at
published code can often be left with many questions about why certain
choices were made. Developing an approach to reveal some of this process
without requiring a reader to essentially reconstruct the analysis
process from the raw data would provide an improved basis for trusting
an analysis result \citep{peng2021reproducible}.

In this paper, we conceptualize these internal expectations and
assumptions as \emph{analysis validation checks}, which allows us to
examine the assumptions made during an analysis and to diagnose
unexpected outcomes. We then introduce a procedure that provides a
quantitative measure of how violations in a tree of analysis checks,
derived from individual checks, will lead to an unexpected result. The
procedure, based on simulations of the original data, calculates the
accuracy and redundancy of the analysis checks. Accuracy is determined
using binary classification metrics, precision and recall, from a logic
regression fit \citep{ruczinski_logic_2003}, while redundancy is
measured using mutual information. The proposed workflow offers a
numerical guarantee that the analysis will produce the expected results,
assuming the assumptions about the data generating mechanism hold.

The rest of the paper is organized as follows:
Section~\ref{sec-lit-review} reviews the concepts of diagnosing
unexpected outcomes and general data quality checks.
Section~\ref{sec-plan} introduces the concept of analysis validation
checks, illustrated with a toy example based on fitness step count data.
Section~\ref{sec-method} describes the procedure that quantifies how
analysis validation checks combined using logical operators can predict
unexpected outcomes in an analysis. Section~\ref{sec-pm10-mortality}
applies this procedure to a larger example that estimates the effect of
particulate matter air pollution on daily mortality.
Section~\ref{sec-discussion} discusses a few key considerations and
Section~\ref{sec-conclusion} concludes the paper.

\section{Related Work}\label{sec-lit-review}

@@

The concept of framing data analysis as a sense-making process was
originally presented by \citep{grolemund_cognitive_2014} based on
seminal work by \citep{wild1999statistical}. Key to any sense-making
process is a model for the world (i.e.~expectations for what we might
observe) and observed data with which we can compare our expectations.
If there is a significant deviation between what we observe and our
expectations, then a data analysis must determine what is causing that
deviation. A naive approach would be to update our model for the world
to match the data, under the assumption that the initial expectation was
incorrect. However, experienced analysts know that the reality can be
more nuanced than that, with errors occurring in data collection or data
processing that can have an impact on final results.

The skill of diagnosing unexpected data analysis results is not one that
has received significant attention in the statistics literature. While
the concept of diagnosis is often embedded in model checking or data
visualization techniques, systematic approaches to identifying the root
cause of an unexpected analysis result are typically not presented
\citep{peng2022perspective}. \citep{peng_diagnosing_2021} proposed a
series of exercises for training students in data analysis to diagnose
different kinds of analysis problems such as coding errors or outliers.
They provide a systematic approach involving working backwards from the
analysis result to identify potential causes. There are parallels here
to the concept of debugging and testing in software engineering
\citep{donoghue2021teaching}. For example, \citep{li2019towards} found
that experienced engineers were generally able to identify problems in
code faster than novices, and that the ability to debug code required
knowledge that cut across different domains.

If it is true that the speed with which data analysts can identify
problems with an analysis is related to their experience working with a
given type of data, then there is perhaps room to improve the analytic
process by externalizing the aspects that an analyst learns through
experience. That way, inexperienced analysts could examine the thought
process of an experienced analyst and learn to identify factors that can
cause unexpected results to occur.

@@

A substantial body of literature has addressed the definition of data
quality \citep[more]{8642813} and has developed frameworks that include
dimensions, attributes, and measures to evaluate and improve data
quality
\citep{cai2015challenges, wang1996beyond, 6204995, woodall2014classification}.
These frameworks are often used in information systems and database
management and support business decision-making in various industries.
For research purposes, high-quality data ensures the credibility of
scientific findings and supports reproducibility and re-usability in
future studies {[}ref{]}. With the growing prevalence of open data in
scientific research, the consumers or users of the data typically are no
longer the producers or collectors of the data who would have the full
knowledge of data in hand, prompting more interest towards data quality
checks in the data analysis process. In R, there are some packages, like
\texttt{skimr} \citep{skimr} and \texttt{dataMaid} \citep{dataMaid},
that provide basic data screening and reporting tools, while another
class of packages, e.g.~\texttt{assertr} \citep{assertr},
\texttt{validate} \citep{validate}, and \texttt{pointblank}
\citep{pointblank} focuses on providing data validation tools, allowing
users to define customized data quality checks based on the
applications.

The literature on data quality typically focuses on the intrinsic or
inherent quality of the data themselves, rather than the data's
relationship to any specific data analysis. So for example, if a column
in a data table is expecting numerical data, but we observe a character
value in one of the entries, then that occurrence would trigger some
sort of data quality check. This type of quality check can be triggered
without any knowledge of what the data will ultimately be used for.
However, for a given analysis, we may require specific aspects of the
data to be true because they affect the result being computed.
Conversely, certain types of poor quality data may have little impact on
the ultimate result of an analysis (e.g.~data that are missing
completely at random). Defining data quality in terms of what may affect
a specific analysis outcome or result has the potential to open new
avenues for defining data checks and for building algorithms for
optimizing the collection of checks defined for a specific analysis.

@@

Expectations represent our understanding of certain aspects of the
analysis and the data, independent of the results of the analysis
itself. When observed outcomes deviate from these expectations, analysts
often revisit the analysis process to identify potential issues, refine
methods, or revise assumptions. Experienced analysts can typically
identify issues quickly and correct them on the spot, but they often do
so without discussing the underlying reasoning, making it harder for
less experienced researchers to learn and master these skills.

Here, we introduce the concept of \textbf{analysis validation checks},
which frame these expectations or assumptions as explicit checks that
return a TRUE or FALSE result given the data analyzed at hand. Inspired
by the concept of data validation checks (\citet{validate}), which are
designed to ensure that datasets meet expected formats and quality,
analysis validation checks reverse the approach: they validate the
assumptions about the data necessary for the analysis to produce the
\emph{expected results}, as defined by the analyst. The focus on
expected results allows the concept of analysis validation checks to
encompass a broad range of checks, such as data quality (i.e.~missing
data, how the data are structured), data distribution and outliers,
bivariate and multivariate relationships between variables, and other
contextual information.

Our proposed analysis validation checks provide insights into an
analyst's thought process and offer the following benefits:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Serve as clear checkpoints to support the replication or application
  of methods to (new) data by programmatically communicating the
  requirements or assumptions made of the data;
\item
  Align assumptions among researchers from different domain backgrounds
  who may have different expectations about the data;
\item
  Improve analysis transparency, reproducibility, and trustworthiness by
  externalizing a key part of the analysis process; and
\item
  Quantify the effectiveness of analysis checks for predicting the
  expected outcome (see Section~\ref{sec-method});
\end{enumerate}

In addition to the above benefits, the development and publication of
analysis checks has the potential to help students, inexperienced
analysts, and junior researchers develop the skills needed to diagnose
unexpected analysis results for a given type of data because the
assumptions made about the data are made transparent. The analysis
checks can serve as a basis for new analysts to have conversations about
the data they are analyzing and to develop a better understanding of the
potential data generation process.

\subsection{A Toy Example}\label{sec-toy}

Consider a 30-day step count experiment in public health. Subjects are
instructed to walk at least 8,000 steps each day, with an expected
average of 9,000 steps, tracked by a step counter app. After 30 days, we
review the data and examine the number of steps taken each day. With
data of this nature, we may expect there to be occasional ``low'' days
due to factors such as forgetting to wear their watch or unfavorable
weather conditions limiting outdoor activities. We may also expect
``high'' days recorded after an outdoor hike or intense workout. Given
the requirements of the study, we expect that for a given subject there
would be no more than five days with step counts falling below the
8,000-step threshold (``invalid days''). If there were more than five
invalid days, the data from that subject would not be usable for the
study.

In this scenario, the outcome is that the number of days with a step
count below 8,000 and the expectation is that this outcome would take a
value no more than 5. If the expectation is violated, we can consider
the potential configurations of the data that could lead to such an
anomaly. To diagnose potential reasons why this outcome expectation
might fail, we can establish a few analysis validation checks in
anticipation of seeing the data. For example, if the average step count
is too low, this may suggest that the subject is incapable of taking the
required number of steps, potentially leading to the unexpected outcome.
Similarly, we can also check the quantile of the step count, if more
than a third of the days fall below 8,000, this could indicate an excess
of low-count days. Additionally, we may may expect the standard
deviation of the step count not to be overly large. These considerations
yield the following three analysis validation checks:

\begin{itemize}
\tightlist
\item
  test1: the test fails if the mean step count is below 8,200
\item
  test2: the test fails if the 30th percentile of the observed step
  counts is below 8,200
\item
  test3: the test fails if the standard deviation of the observed step
  counts exceeds 2,500.
\end{itemize}

The cutoff values chosen for these tests would presumably be chosen
based on prior experience with these kinds of data, but could also be
optimized using the method presented in the next section.

@@

To simulate this data, three normal distributions are used for the daily
step counts: \(\mathcal{N}(4000, 200)\) for low days,
\(\mathcal{N}(12000, 200)\) for high days, and
\(\mathcal{N}(9000, 300)\) for typical days. The number of low and high
days can be simulated from a Poisson distribution with \(\lambda = 4\).
Figure~\ref{fig-step-count} displays the number of days with fewer than
8,000 steps across 300 simulated 30-day periods.

@@

While some checks may be crucial and directly indicate an unexpected
outcome, others may be tangential to the problem at hand and not
indicate a root cause of an unexpected outcome. In this section, we
propose a procedure to measure the effectiveness of checks that, when
combined using logical operators, contribute to an unexpected outcome. A
small set of independent checks is considered effective if it translates
to unexpected outcomes.

The approach relies on the use of simulated datasets that are generated
based on the analyst's knowledge and assumptions about the data
generation mechanism. Datasets are simulated to have the same structure
and characteristics of the observed data that will be analyzed at some
point in the future. For each simulated dataset, we can apply a
collection of analysis validation checks to the dataset and record which
ones were TRUE and which were FALSE. We can also compute the outcome of
the analysis to see whether the outcome is unexpected in a given
simulated datsets. We can then generate many datasets, each time
applying the analysis validation checks and computing the outcome. After
generating many datasets, we can relate the patterns in the analysis
validation checks to the likely that the outcome will be unexpected in a
given dataset. Figure~\ref{fig-metric-calc} provides an overview of the
process.

@@

From the simulated data, the accuracy branch refers to a set of checks'
ability to accurately detect unexpected outcomes while minimizing false
positives and false negatives. While a false positive can raise caution
or skepticism on the data, the presence of both false positives and
false negatives suggest that the checks are overly sensitive or lack
sensitivity to unexpected outcomes.

To incorporate the effect of multiple checks on the outcome, a logic
regression model \citep{ruczinski_logic_2003} is fitted to the analysis
validation checks using the indicator of an unexpected result as the
outcome in the model. Originally developed for SNP microarray data,
logic regression constructs Boolean combinations of binary variables, in
a tree structure, to predict both binary and numerical outcomes.
Compared to other tree-based methods for binary-binary prediction, the
Boolean combinations from the logic regression model produce a tree
structure that can be directly interpreted as the possible combination
of checks leading to an unexpected outcome, without the need to invert
the tree as required in classic tree-based recursive partitioning
methods. The logic regression model is then used to predict the analysis
result based on the values of checks, and the prediction is compared to
the actual analysis result in order to calculate the precision and
recall of the checks.

While tests may score high on accuracy, they may be less effective at
explaining the various reasons behind unexpected results. This could
happen if, for example, a set of tests are all tangentially related to
the cause of the unexpected results, but none addresses the root cause.
It may also occur if the tests are highly correlated with one another,
leading to redundancy.

@@

To combine precision, recall, and independence into a single metric, we
can combine the three scores using the arithmetic mean, harmonic mean,
or quadratic mean. The differences among these means are minimal when
the three metrics are similar. However, as the differences among the
metrics increases, the harmonic mean tends to produce the smallest
overall score, as it penalizes low values, while the quadratic mean
tends to produce the largest score by rewarding higher values more. For
simple interpretation of the score, the arithmetic mean is preferred,
while in applications where the difference between precision, recall,
and independence need to be penalized or rewarded more, the harmonic and
quadratic mean should be considered.

\subsection{Toy Example Revisited}\label{toy-example-revisited}

Returning to the step count example introduced in Section~\ref{sec-toy},
the logic regression model is fitted to the three analysis checks
described previously to generate the prediction of the unexpected
outcome, which is whether the number of days with fewer than 8,000 steps
is greater than 5. The predictions from the logic regression model can
be compared with the simulated true outcome for calculating the
precision and recall metrics. Figure~\ref{fig-logic-reg} shows the
best-fitting logic regression model, which is a combination of test1
(``mean step count is below 8,200'') and test3 (``standard deviation
exceeds 2,500'') with an OR operator. In other words, if either test1 or
test3 is true, then we would predict an unexpected outcome in the
analysis (i.e.~the number of days with fewer than 8,000 steps is greater
than 5).

@@

Table~\ref{tbl-logic-reg} presents the calculated precision, recall, and
independence for the three individual tests and the combined test rule
(test1 OR test3) from the logic regression. We also include the metric
calculated from fitting a regression tree model to the data to compare
the performance of the logic regression model. The harmonic and
arithmetic means are included to combine the three measures. The results
show that the two tests produced by the logic regression can accurately
predict 83.6\% cases of all \emph{actual unexpected results} in the
simulation data. Furthermore, 82.1\% of all \emph{predicted unexpected
results} were in fact observed to be unexpected.

@@

For comparison, the regression tree produces a similar prediction to the
logic regression, by first splitting on test1 and then splitting on
test3, and results in the same accuracy and overall score as the logic
regression model. However, we argue that the logic regression tree shown
in Figure~\ref{fig-logic-reg} is more interpretable for our purposes
because it provides a direct representation of which combinations of
analysis checks lead to unexpected outcomes. The logic regression tree
is also directly comparable to other diagnostic techniques, which we
discuss further in Section~\ref{sec-discussion}.

@@

In the study of the health effects of outdoor air pollution, one area of
interest is the association between short-term, day-to-day changes in
particulate matter air pollution and daily mortality counts. Substantial
work has been done to study this question and to date, there appears to
be strong evidence of an association between particulate matter less
than 10 \(\mu\)g/m\(^3\) in aerodynamic diameter (PM10) and daily
mortality from all non-accidental causes \citep{samet2000fine}. For our
second example, we use the problem of studying PM10 and mortality along
with data from the National Morbidity, Mortality, and Air Pollution
Study (NMMAPS) to demonstrate how our analysis validation checks
described in Section~\ref{sec-method} can be applied.

The typical approach to studying the association between PM10 and
mortality is to apply a generalized linear model with a Poisson link to
relate daily mortality counts to daily measures of PM10. Based on
previous work and the range of effect sizes published in the literature,
an analyst might expect the coefficient for PM10 in this GLM to lie
between {[}0, 0.005{]}, after adjusting for daily temperature
\citep{samet200fine, welty2005acute}. Observing an estimated coefficient
outside of this interval would be highly unusual and would warrant a
serious re-examination of the analysis process. Therefore, our
unexpected outcome in this analysis will be the binary indicator of
whether the estimated coefficient for PM10 in the GLM lies outside of
the interval {[}0, 0.005{]}. In addition to providing a more substantial
problem for our methods, this example also demonstrates how the
procedure presented in Section~\ref{sec-method} can be used to select
cutoff values in the analysis checks to diagnose an unexpected PM10
coefficient from the generalized linear model.

This PM10 coefficient expectation can be framed as an analysis check
that fails, labelled as 1, if the estimate of the PM10 coefficient
(adjusted for temperature) is outside the range {[}0, 0.005{]}, and 0 if
it is within. Multiple factors can affect the estimated PM10
coefficient, such as the sample size, the strength of the correlation
between mortality and PM10, and the strength of the correlation between
mortality and temperature. Analysts may expect a reasonable sample size
to ensure the reliability of the coefficient estimate. Outliers in the
three variables can also leverage the coefficient. While these are
possible factors that could affect the analysis result, it is not clear
what the cutoff values for these checks should be to determine a
failure. Here we consider a list of checks in Table~\ref{tbl-checks}
with varied cutoff values for each:

@@

\subsection{Data Simulation}\label{data-simulation}

To generate replicates of the dataset, we first generate the correlation
matrix of the three variables (PM10, mortality, and temperature) in a
grid and then use a Gaussian copula to generate a multivariate normal
distribution based on the specified correlation matrix and sample size.
The multivariate normal distribution is transformed using the normal CDF
before the inverse CDF of the assumed distributions of the three
variables is applied. To determine the appropriate distribution of each
variable, various distributions are fitted and compared. This includes
poisson and negative binomial for mortality; gamma, log-normal,
exponential, weibull, and normal for PM10 and temperature; and beta for
PM10 after rescaling the data to \([0,1]\).

To ensure a reasonable likeness to data that might be used in such an
analysis, we use characteristics of the observed dataset to refine our
simulations. AIC is used to determine the best distribution fit for each
variable with the QQ-plot presented in Figure~\ref{fig-dist-fit} to
evaluate the fit. AIC suggests a negative binomial distribution for
mortality, a beta distribution for PM10 (multiple by 100 to recover the
original scale), and a Weibull distribution for temperature. To include
the potential effect of outliers, we add a single outlier to the data
for both the mortality and PM10 variables.

@@

A logic regression is fitted using all variations of the tests in
Table~\ref{tbl-checks} to predict whether the PM10 coefficient is
unexpected. Figure~\ref{fig-linear-reg-tree} shows the optimal logic
regression tree from the fitted model. Precision, recall, and
independence score, along with their harmonic and arithmetic mean are
calculated in Table~\ref{tbl-linear-reg}.

@@

As indicated in Figure~\ref{fig-linear-reg-tree}, the logic regression
model picks up the following cutoff values for each type of check:

\begin{itemize}
\tightlist
\item
  sample size \emph{larger than} 200
\item
  mortality-temperature correlation greater than \(-0.35\)
\item
  mortality-PM10 correlation less than \(-0.04\)
\item
  mortality data contains outliers that are detected by the univariate
  outlier detection
\end{itemize}

@@

As shown in Figure~\ref{fig-linear-reg-tree}, there is no single
analysis check in the tree that predicts an unexpected outcome. rather
at least three checks in the tree must be TRUE in order for the model to
predict an unexpected outcome. Given the high independence of the tests
(Table~\ref{tbl-linear-reg}), this suggests that unexpected results are
only likely after multiple anomalies are observed in the data.
